An Analysis of Performance Bottlenecks in MRI Pre-Processing

### 2024-03-28 paper/paper.tex
Additional comments

\section{Materials \& Methods}
\subsection{Pipelines}
We study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow. First, we profile the entire workflow to identify the coarse grain performance bottleneck to obtain a coarse view of the workflow's performance. Following, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}. We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset. Moreover, the statistical analysis of the fMRIPrep workflow are not computationally expensive and are not the focus of this paper.

We use the default parameters for fMRIPrep and most of the sub-workflows. However, for FreeSurfer recon-all, we mimic the fMRIPrep specific 3 stages workflow.

We profile both the single and double precision version for ANTs registrationSyN and ANTs brainExtraction. ANTs uses ITK templates to implement the single precision alternative.

\subsection{Data}
To ensure the performance profiles of the applications derived from our work are inclusive of different populations, we used data with a wide range of age and equal distribution of sex (biological). Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.

We used the the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}. The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts. The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two session: eyes open and eyes closed. The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session. We profiled the application using both cohort on the eyes open session, unless explicitly specified.

\subsection{Profiling}
Profiling MRI pre-processing pipelines present a set of challenges. The neuroimaging community develops pipelines using several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Matlab}, and \textit{Python}, with several pipelines utilizing a combination of multiple languages. Pipelines are generally complex and computationally expensive.

The Intel VTune profiler solves these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level. However, common profiling challenges remain. Profilers require source code to be compiled with Debug Symbols to report useful information about the different functions and modules. Moreover, different input data can substantially vary the analysis due to conditional branching in the application and convergence thresholds. We discuss our approach to these challenges in the next section.
\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for references}

First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image. For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{alg:vtune_example} illustrate the profiling of a pipeline by mounting the VTune binary to the Apptainer image at execution.

@@

Second, we aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile. The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes. We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature~\cite{Kukunas2015-jd}. Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the application. This heuristic allows us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.

Last, we profile the application using two threading configuration: single-threaded and 32 threads. On the one hand, the single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling. 
On the other hand, we use all available CPU cores from a compute node (i.e. 32) to study the multi-threading performance of the application. This is a more realistic usage for the user of the application.

@@

The data generated by the VTune profiler is written to a shared file system, for administrative reasons. Theoretically, this increase the profiling overhead. However, in practice there is no impact since the data generated by the VTune profiler is very small and is not written on the same file system as the data from applications.

\subsection{Metrics Definition}
We define the makespan as the elapsed time from the start to the end of the application. The CPU time refers to the total time spent by the CPU, across all core, to execute the application. VTune provides several metrics to characterize the performance of an application. We describe the memory metrics used in this paper.

\begin{figure*}[h!]
	\noindent
	\begin{minipage}
		{0.5\textwidth}
		\begin{align*}
			C & = CPU\_CLK\_UNHALTED.THREAD                \\
			S & = CYCLE\_ACTIVITY.STALLS\_LDM\_PENDING     \\
			O & = CYCLE\_ACTIVITY.STALLS\_L1D\_PENDING     \\
			T & = CYCLE\_ACTIVITY.STALLS\_L2\_PENDING      \\
			W & = MEM\_L3\_WEIGHT                          \\
			H & = MEM\_LOAD\_UOPS\_RETIRED.LLC\_HIT        \\
			R & = MEM\_LOAD\_UOPS\_MISC\_RETIRED.LLC\_MISS \\
			M & = \frac{W \times R}{H + R}                 \\
			N & = \frac{H}{H + W \times R}                 
		\end{align*}
	\end{minipage}
	\begin{minipage}
		{0.5\textwidth}
		\begin{align}
			\%~Memory~Bound & = \frac{S}{C} \times 100          \\
			\%~L1~Bound     & = \frac{S - O}{C} \times 100      \\
			\%~L2~Bound     & = \frac{O - T}{C} \times 100      \\
			\%~L3~Bound     & = \frac{T \times N}{C} \times 100 \\
			\%~DRAM~Bound   & = \frac{T \times M}{C} \times 100 
		\end{align}
	\end{minipage}
	\caption{Intel performance monitoring events and derived memory metrics}
	\label{fig:memory-metrics}
\end{figure*}

The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load. L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache. Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively. The store bound occurs when a store operation stall the CPU. Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}. We use the convention from~\cite{Kukunas2015-jd} for the equations.

The difference between the memory bound value and the sum of the other metrics represent the percentage of CPU starvation while waiting from data absent from all cache level; i.e. the data is fetched from disk.

\subsection{Infrastructure}
For our experiments, we used the \textit{Slashbin cluster} at Concordia University. The compute nodes are configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), Rocky~Linux~8.9 and Linux kernel \textit{4.18.0-477.10.1.el8\_lustre.x86\_64}.

\section{Results}
In this section we present the aggregated profiling data obtained with VTune. First, we present a performance overview of the pipelines. We then dive into more detailed analysis observed from the hotspots of the pipelines. Unless specified explicitly, we only present the profile data from using 1-thread since the results from 32-threads are similar.

\subsection{Long-tail Distribution}
Figure~\ref{fig:long-tail-distribution} shows that 80\% of the total CPU time is spent in less than 1.5\% of the function across all pipelines. While consistent with the Pareto principle, the distribution of the functions' CPU time does not follow a pareto distribution. We plotted the distribution of the functions' CPU time on a log-log scale and found it to be closer to a Zipf's law distribution with exponent $\alpha=1.8$.

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{figures/global-longtail.png}
	\caption{Distribution of the functions' CPU time. The left y-axis shows the average total CPU Time spent in a function, while the right y-axis show the cumulative CPU time percentage. The x-axis percentage of functions ordered by decreasing CPU time. The data includes all functions from all pipelines.}
	\label{fig:long-tail-distribution}
\end{figure}

\subsection{Makespan: A performance overview}
Figure~\ref{fig:makespan-1thread} depicts the total time elapsed from the start to the end of each pipeline; i.e. makespan. There is a significant difference between the makespan of different applications. This is expected as the pipelines perform different tasks. The standard deviation is nearly null for all pipelines, except for FreeSurfer recon-all which as a small variation. Overall, this shows that the input data does not substantially affect the makespan of those pipelines.

@@

\subsection{Impact of Memory Bounded Functions}
Table~\ref{extab:memory-bound} shows that fetching data from disk is the main source of cycles where the CPU is starved. Following, The L1 bound has the highest cache-level contribution to the memory bound of the pipelines. ANTS pipelines have the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT. L2 and L3 bound is low for all pipeline, although highest for FreeSurfer recon-all. The DRAM bound is high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTS brainExtraction in single and double precision. Store bound is low for all pipelines. We note an interesting relationship between the low L1 bound and the high DRAM bound of FreeSurfer recon-all and FSL FLIRT, compared to the other pipelines.

@@

\subsection{Main Bottleneck: Linear Interpolation}
Table~\ref{extab:interpolation} shows that interpolation (or convolution) is a critical part for all pipelines, except FreeSurfer recon-all. For those pipelines, interpolation contributes between 32.20\% to 62.70\% of the total CPU time. Overall, the number of functions using interpolation is low with less than 20 in each applications. In other words, few interpolation functions are contribute to a large portion of the CPU time.
% There seems to be few  the stages from recon-all: https://surfer.nmr.mgh.harvard.edu/fswiki/recon-all (see section "Manual-Intervention Workflow Directives": stages 1-31)

@@

The number of iteration between both version of ANTS registrationSyN is similar. In fact, for the latest two stages, they were the same across all but one execution. However, The single precision version of ANTS registrationSyN took longer per iteration than the double precision version (Figure~\ref{fig:mean-time-per-iteration-ants}). Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration.

@@

\subsection{FreeSurfer: Thread-Synchronization}
Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant different in the pipeline profile between single-threaded and multi-threaded execution. The multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime. FreeSurfer uses the multi-processing API provided by OpenMP. Moreover, the y-axis scale is multi folds larger than for the single-threaded executions. This indicates that the multi-threaded execution is significantly slower than the single-threaded execution.

@@

While the makespan for FreeSurfer recon-all decrease with an increase in the number of threads, the benefits are limited (Figure~\ref{fig:freesurfer-threading}). The parallel efficiency decreases from 66\% with two threads down to 5\% with 32 threads. While the Amdhals' law plays an impact on the maximum parallel efficiency of the pipeline, Figure~\ref{fig:hotspots-freesurfer-reconall} indicates that the thread Synchronization has a larger impact.

@@

\section{Discussion}
\subsection{Few functions account for most of the CPU time}
Our initial assumption was that the distribution of CPU time would follow a Pareto distribution. Thus, 20\% of the functions would contribute to 80\% of the total CPU time. However, Figure~\ref{fig:long-tail-distribution} shows that the distribution of CPU time is closer to a Zipf's law with exponent $\alpha=1.8$. This means that the CPU time of a function is approximately inversely proportional to its rank. This is consistent with the Pareto's principle, but is more extreme. Therefore, future efforts to optimize the main bottlenecks of MRI pre-processing pipelines could lead to enormous performance improvements.

\subsection{Wasted CPU cycles due to Memory Bound}
Overall, our results show a high percent of memory bound across all pipelines.
Optimizing the data access of the application could reduce the number of data load. This would potentially have a large impact on the performance of the pipelines. Alternatively, the use of reduced precision could reduce the memory footprint, thus the memory bound. To reduce the memory bound impact of fetching data from disk, it could be possible to convert and store the data to a lower precision before pre-processing. In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.

@@

There is an extensive literature on interpolation. Unfortunately, large amount of optimization techniques are not applicable to MRI pre-processing pipelines, because of the rotation during registration which requires dynamic kernels for interpolation. Moreover, the high-resolution of the data and iterative refinement of the registration require a large number of interpolations to be performed.

The authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation. The review algorithm trade-off accuracy for a lower computational cost. We think it would be interesting for future work to explore these different alternatives to accelerate interpolation within the pipelines. However, special attention will be required to balance the accuracy and the performance of the algorithm.

@@

Improvements to interpolation in 3D MRI imaging would have a large impact. It would also benefit other fields of studies using interpolation for high-resolution imaging with iterative refinement, such as climate simulation, astronomical imaging, or 3D image reconstruction.

\subsection{ANTS: A tale of precision}
From our results, we observe a slow-down for the single precision version of ANTs pipelines. A single iteration takes longer to execute when using single-precision compared to double precision. After a further analysis of the ANTs registrationSyN single-precision pipeline, we found that the function\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} contributing most of the CPU time does computations in double precision. Therefore, the pipeline wastes CPU cycles to convert the input data to double in ITK, and potentially to cast to back to float in ANTS. We think this bug is the cause of the slow-down. We tried recompiling a version of ITK to fix this issue, but we were unsuccessful due to the size and complexity of the code base. This example shows that the use of reduced precision is not trivial and require a deep understanding of the code base. Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline.

\subsection{OpenMP bottleneck in FreeSurfer recon-all}
In our experiments, FreeSurfer recon-all showed a significant decrease in parallel efficiency with an increase in the number of threads. While the Amdhals' law plays an impact on the maximum efficiency, the OpenMP thread synchronization played the largest impact in our results. The codebase use 83 \textit{static} scheduling policy out of 92 total. \textit{Static} scheduling assign chunks to threads in round-robin fashion. This could lead to a load imbalance between the threads, and thus a decrease in parallel efficiency. We supposed that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads. Unfortunately, we failed to recompile FreeSurfer by naively changing the \textit{static} OpenMP scheduling to \textit{dynamic}. This remain challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs.

@@

\subsection{Limitations}
The dataset we chose contains wide range of age and equal distribution of sex. However, it only contains healthy subjects. Therefore, the performance bottleneck we found might not be applicable to data pre-processing of pathological patients. Moreover, the dataset is from a single site. This could also introduce bias from the scanner model and data acquisition protocol used. Future work should include data from multiple sites to ensure the generalization of the performance bottleneck we found.

Our analysis omits the space or time of execution for function calls. This information could provide additional insight for optimization. For example, spatial and temporal dimensions were shown to be important in the field of reduced precision. Future work could extent the depth of the analysis by including this information.
\MD{TODO Cite paper on spatial/temporal for RP (ask yohan for references)}

\section{Conclusion}
In this paper, we profiled several MRI pre-processing applications: ANTS brainExtraction, ANTS registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all, and fMRIPrep. We confirmed our hypothesis that few functions contribute to a large amount of CPU time. We found that most applications suffered from memory bound. Linear interpolation was the primary bottleneck for these applications. We discovered a bug in ITK which lead the implementation of ANTS registration in single-precision to use float for its computation. We also discovered a potential bug in FreeSurfer recon-all which limits the benefit from multi-threading with OpenMP. Last, we discuss the challenge of profiling long running applications with VTune, due to computational resource requirements to obtain results.

>>>>

\section{Materials \& Methods}
\subsection{Pipelines}
We study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow. First, we profile the entire workflow to obtain a coarse view of the workflow's performance. Furthermore, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}. We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset \TG{add a few words to explain why}. \TG{Information is missing to explain what is the statistical analysis in fMRIPrep. Likewise, it's unclear what ANTS registration, FSL Flirt, etc are and how they are combined in fMRIPrep. This section should start with a higher-level overview of fMRIPrep. You should also explain that fMRIPrep's building blocks are widely used in neuroimaging.} Moreover, the statistical analysis of the fMRIPrep workflow is not computationally expensive and is not the focus of this paper.

We use the default parameters for fMRIPrep and most of the sub-workflows. However, for FreeSurfer recon-all, we mimic the fMRIPrep specific 3 stages workflow \TG{This is cryptic, it needs more context.}.

We profile both the single and double precision version for ANTs registrationSyN and ANTs brainExtraction. ANTs uses ITK templates to implement the single precision alternative. \TG{You should re-order this paragraph as follows: (1) ANTs registration has two implementations, a double precision one and a single precision one. (2) This is actually coming from ITK. (3) We benchmark both implementations.}

\subsection{Data}
To ensure that our performance profiles are inclusive of different populations, we used data with a wide range of age and equal distribution of sex. We used the the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}. The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts. The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two session: eyes open and eyes closed. The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session. We profiled the application using both cohorts and the eyes open session, unless explicitly specified \TG{is it ``explicitly specified'' anywhere? If not, remove this.}.

\subsection{Profiling}
Profiling MRI pre-processing pipelines \TG{the terms pipelines, workflfows, applications are used interchangeably, let's harmonize and use only 1 term (I don't mind which one)} raises a few challenges. Neuroimaging pipelines use several programming languages including \textit{C}, \textit{C++}, \textit{Fortran}, \textit{Matlab}, and \textit{Python}, with several pipelines utilizing a combination of multiple languages. Pipelines are generally complex and computationally expensive.

The Intel VTune profiler addresses these challenges by offering multi-language support, low performance overhead, and information of functions at runtime level. However, common profiling challenges remain. Profilers require source code to be compiled with debug symbols to report interpretable information about the different functions and modules. Moreover, different input data can substantially impact the performance results due to conditional branching in the application and convergence thresholds. We discuss our approach to these challenges in the next section.
\MD{TODO Cite effect from varying input data on numerical stability and reduced precision. Ask yohan for references}

First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image. For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{alg:vtune_example} \TG{Could this be called a listing rather than an algorithm?} illustrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.

@@

Second, we aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile. The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes. We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature~\cite{Kukunas2015-jd}. Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the application \TG{This discussion on Pareto principle should be moved to the results section}. This heuristic allows us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.

Last, we profile the application using two threading configuration: single-threaded and 32 threads. The single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling. 
The 32-thread approach uses all available CPU cores from a compute node of our infrastructure to study the multi-threading performance of the application. This is a more realistic usage for the user of the application.

@@

The data generated by the VTune profiler is written to a shared file system, for administrative reasons. In principle, this could increase the profiling overhead. However, in practice there is no impact since the data generated by the VTune profiler is very small and is not written on the same file system as data produced by the applications.

\subsection{Metrics Definition}
VTune provides several metrics to characterize the performance of an application.  The makespan is the elapsed time from the start to the end of the application. The CPU time refers to the total time spent by the CPU, across all cores, to execute the application. 
\begin{figure*}[h!]
	\noindent
	\begin{minipage}
		{0.5\textwidth}
		\begin{align*}
			C & = CPU\_CLK\_UNHALTED.THREAD                \\
			S & = CYCLE\_ACTIVITY.STALLS\_LDM\_PENDING     \\
			O & = CYCLE\_ACTIVITY.STALLS\_L1D\_PENDING     \\
			T & = CYCLE\_ACTIVITY.STALLS\_L2\_PENDING      \\
			W & = MEM\_L3\_WEIGHT                          \\
			H & = MEM\_LOAD\_UOPS\_RETIRED.LLC\_HIT        \\
			R & = MEM\_LOAD\_UOPS\_MISC\_RETIRED.LLC\_MISS \\
			M & = \frac{W \times R}{H + R}                 \\
			N & = \frac{H}{H + W \times R}                 
		\end{align*}
	\end{minipage}
	\begin{minipage}
		{0.5\textwidth}
		\begin{align}
			\%~Memory~Bound & = \frac{S}{C} \times 100          \\
			\%~L1~Bound     & = \frac{S - O}{C} \times 100      \\
			\%~L2~Bound     & = \frac{O - T}{C} \times 100      \\
			\%~L3~Bound     & = \frac{T \times N}{C} \times 100 \\
			\%~DRAM~Bound   & = \frac{T \times M}{C} \times 100 
		\end{align}
	\end{minipage}
	\caption{Intel performance monitoring events and derived memory metrics}
	\label{fig:memory-metrics}
\end{figure*}	
The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load. L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache. Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively. The store bound occurs when a store operation stall the CPU. Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc} \TG{This figure should go to appendix, it's not critical to the understanding of the paper}. We use the convention from~\cite{Kukunas2015-jd} for the equations.

The difference between the memory bound value and the sum of the other metrics represent the percentage of CPU starvation while waiting from data absent from all cache level, that is, when the data is being fetched from disk.

\subsection{Infrastructure}
We used the \textit{Slashbin cluster} at Concordia University. The compute nodes are configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), Rocky~Linux~8.9 and Linux kernel \textit{4.18.0-477.10.1.el8\_lustre.x86\_64}.

\section{Results}
In this section we present the aggregated profiling data obtained with VTune. First, we present a performance overview of the pipelines. We then analysis specific hotspots in the pipelines. Unless specified explicitly, we only present the profile data from using 1-thread since the results from 32-threads are similar.

\subsection{CPU time distribution across functions was long tailed}
Figure~\ref{fig:long-tail-distribution} shows that 80\% of the total CPU time was spent in less than 1.5\% of the function across all pipelines. While consistent with the Pareto principle, the distribution of the functions' CPU time does not follow a Pareto distribution. We plotted the distribution of the functions' CPU time on a log-log scale and found it to be closer \TG{``closer'' is too vague: is it well approximated by a Zipf's law? Can you provide a p-value of this claim?} to a Zipf's law distribution with exponent $\alpha=1.8$. \TG{Add a few sentences to explain why this is interesting.}

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{figures/global-longtail.png}
	\caption{Distribution of the functions' CPU time. The left y-axis shows the average total CPU Time spent in a function, while the right y-axis show the cumulative CPU time percentage. The x-axis percentage of functions ordered by decreasing CPU time. The data includes all functions from all pipelines. }
	\label{fig:long-tail-distribution}
\end{figure}

\subsection{Makespan: A performance overview}
Figure~\ref{fig:makespan-1thread} depicts the total time elapsed from the start to the end of each pipeline; i.e. makespan. There is a significant difference between the makespan of different applications. This is expected as the pipelines perform different tasks. The standard deviation is nearly null for all pipelines, except for FreeSurfer recon-all which as a small variation. Overall, this shows that the input data does not substantially affect the makespan of those pipelines. \TG{not sure what to make from this result, I don't find it particularly noticeable, I think I would omit it.}

@@

\subsection{Impact of Memory Bounded Functions}
Table~\ref{extab:memory-bound} shows that fetching data from disk is the main source of cycles where the CPU is starved \TG{I don't see disk in table 1}. Following, the L1 bound has the highest cache-level contribution to the memory bound of the pipelines. ANTS pipelines have the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT. L2 and L3 bound is low for all pipeline, although highest for FreeSurfer recon-all. The DRAM bound is high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTS brainExtraction in single and double precision. Store bound is low for all pipelines. We note an interesting relationship between the low L1 bound and the high DRAM bound of FreeSurfer recon-all and FSL FLIRT, compared to the other pipelines. \TG{Add discussion: what is interesting about this result?}

@@

\subsection{Main Bottleneck: Linear Interpolation}
Table~\ref{extab:interpolation} shows that interpolation (or convolution) is a critical part for all pipelines, except FreeSurfer recon-all. For those pipelines, interpolation contributes between 32.20\% to 62.70\% of the total CPU time. Overall, the number of functions using interpolation is low with less than 20 in each applications. In other words, few interpolation functions are contribute to a large portion of the CPU time. \TG{This should be moved up in the results, I think right after Figure 2, as this is one of the most interesting results.} \TG{Add a few sentences to explain why this is interesting.}
% There seems to be few  the stages from recon-all: https://surfer.nmr.mgh.harvard.edu/fswiki/recon-all (see section "Manual-Intervention Workflow Directives": stages 1-31)

@@

The number of iteration between both version of ANTS registrationSyN is similar. In fact, for the latest two stages, they were the same across all but one execution. However, The single precision version of ANTS registrationSyN took longer per iteration than the double precision version (Figure~\ref{fig:mean-time-per-iteration-ants}). Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration. \TG{This is too inconclusive, you should give all the details here.}

@@

\subsection{FreeSurfer: Thread-Synchronization}
Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant difference in the pipeline profile between single-threaded and multi-threaded execution. In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the application, accounting for at least 76\% of the application runtime. FreeSurfer uses the multi-processing API provided by OpenMP. Moreover, the y-axis scale is multi folds larger than for the single-threaded executions. This indicates that the multi-threaded execution is significantly slower than the single-threaded execution.

@@

While the makespan for FreeSurfer recon-all decreases with an increase in the number of threads, the benefits are limited (Figure~\ref{fig:freesurfer-threading}). The parallel efficiency decreases from 66\% with two threads down to 5\% with 32 threads. While Amdhals' law plays an impact on the maximum parallel efficiency of the pipeline, Figure~\ref{fig:hotspots-freesurfer-reconall} indicates that the thread Synchronization has a larger impact.

@@

\section{Discussion}

\TG{I think the discussion should be merged in the results. I know it's unconventional but I think it would make the paper easier to read and more interestingly. You could add a short conclusion that summarizes the main take-aways from the results.}

\subsection{Few functions account for most of the CPU time}
Our initial assumption was that the distribution of CPU time would follow a Pareto distribution. Thus, 20\% of the functions would contribute to 80\% of the total CPU time. However, Figure~\ref{fig:long-tail-distribution} shows \TG{Figure 2 doesn't ``show'' that. Rephrase.} that the distribution of CPU time is closer to a Zipf's law with exponent $\alpha=1.8$. This means that the CPU time of a function is approximately inversely proportional to its rank. This is consistent with the Pareto's principle, but is more extreme. Therefore, future efforts to optimize the main bottlenecks of MRI pre-processing pipelines could lead to enormous performance improvements.

\subsection{Wasted CPU cycles due to Memory Bound}
Overall, our results show a high percent of memory bound across all pipelines.
Optimizing the data access of the application could reduce the number of data load. This would potentially have a large impact on the performance of the pipelines. Alternatively, the use of reduced precision could reduce the memory footprint, thus the memory bound. To reduce the memory bound impact of fetching data from disk, it could be possible to convert and store the data to a lower precision before pre-processing. In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.
\TG{You could cite some of Valerie's works in this direction.}

@@

There is an extensive literature on interpolation \TG{cite it}. Unfortunately, large amount of optimization techniques are not applicable to MRI pre-processing pipelines, because of the rotation during registration \TG{explain how interpolation is related to registration. In your presentation of fmriprep, explain what registration is and the role it plays in the pipelines.} \TG{this is a bit cryptic and I don't think you want to state that here. You may want to say that it would be interesting to further investigate these techniques in the future.} which requires dynamic kernels for interpolation. Moreover, the high-resolution of the data and iterative refinement of the registration require a large number of interpolations to be performed.

The authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation. They review algorithm trade-off accuracy for a lower computational cost. We think it would be interesting for future work to explore these different alternatives to accelerate interpolation within the pipelines. However, special attention will be required to balance the accuracy and the performance of the algorithm.

@@

Improvements to interpolation in 3D MRI imaging would have a large impact. It would also benefit other fields of studies using interpolation for high-resolution imaging with iterative refinement, such as climate simulation, astronomical imaging, or 3D image reconstruction. \TG{You could summarize the previous 3 paragraphs, I find them a bit verbose.}

\subsection{ANTS: A tale of precision}
From our results, we observe a slow-down for the single precision version of ANTs pipelines. A single iteration takes longer to execute when using single-precision compared to double precision. After a further analysis of the ANTs registrationSyN single-precision pipeline, we found that the function\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} contributing most of the CPU time does computations in double precision. Therefore, the pipeline wastes CPU cycles to convert the input data to double in ITK, and potentially to cast to back to float in ANTS. We think this bug is the cause of the slow-down. We tried recompiling a version of ITK to fix this issue, but we were unsuccessful due to the size and complexity of the code base. This example shows that the use of reduced precision is not trivial and require a deep understanding of the code base. Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline. \TG{Can you report the bug and add a link to the issue here?}

\subsection{OpenMP bottleneck in FreeSurfer recon-all}
In our experiments, FreeSurfer recon-all showed a significant decrease in parallel efficiency with an increase in the number of threads. While Amdhals' law plays an impact on the maximum efficiency, the OpenMP thread synchronization played the largest impact in our results. The codebase uses 83 \textit{static} scheduling policy out of 92 total \TG{explain what these numbers represent}. \textit{Static} scheduling assign chunks to threads in round-robin fashion. This could lead to a load imbalance between the threads, and thus a decrease in parallel efficiency. We speculated that \textit{dynamic} scheduling could improve the parallel efficiency, by assigning chunks based on the current load of the threads. Unfortunately, we failed to recompile FreeSurfer by naively changing the \textit{static} OpenMP scheduling to \textit{dynamic}. This remains challenging as the codebase is large and complex, and the change in scheduling policy could lead to unexpected bugs. \TG{Could you file a bug report in freesurfer on this issue and refer to it here?}

@@

\subsection{Limitations}
\TG{You should frame the limitations as future work in the conclusion. These are not really limitations.}
The dataset we chose contains wide range of age and equal distribution of sex. However, it only contains healthy subjects. Therefore, the performance bottleneck we found might not be applicable to data pre-processing of pathological patients. Moreover, the dataset is from a single site. This could also introduce bias from the scanner model and data acquisition protocol used. Future work should include data from multiple sites to ensure the generalization of the performance bottleneck we found. \TG{Is there evidence showing that this would provide additional insights on the benchmark? I think it might be interesting to study subjects that fail QC but I'm not sure about other populations.}

Our analysis omits the space or time of execution for function calls \TG{I don't understand what this means}. This information could provide additional insight for optimization. For example, spatial and temporal dimensions were shown to be important in the field of reduced precision. Future work could extent the depth of the analysis by including this information.
\MD{TODO Cite paper on spatial/temporal for RP (ask yohan for references)}

\section{Conclusion}
In this paper, we profiled several MRI pre-processing applications: ANTS brainExtraction, ANTS registration, FSL FAST, FSL FLIRT, FSL MCFLIRT, FreeSurfer recon-all, and fMRIPrep. We confirmed our hypothesis \TG{don't frame it as a hypothesis, it is an observation} that few functions contribute to a large amount of CPU time \TG{, which [explain why it is interesting to know]}. We found that most applications suffered from memory bound. Linear interpolation was the primary bottleneck for these applications. We discovered a bug in ITK which lead the implementation of ANTS registration in single-precision to use float for its computation. We also discovered a potential bug in FreeSurfer recon-all which limits the benefit from multi-threading with OpenMP. Last, we discuss the challenge of profiling long running applications with VTune, due to computational resource requirements to obtain results.
